### 2016-09-15 revision/abstract.tex
changes for reviewer 1

We design algorithms for online linear optimization that have optimal regret
and at the same time do not need to know any upper or lower bounds on the norm
of the loss vectors.  We achieve adaptiveness to the norms of the loss vectors
by scale invariance, i.e., our algorithms make exactly the same decisions if
the sequence of loss vectors is multiplied by any positive constant.  One of
our algorithms works for any decision set, bounded or unbounded.  For unbounded
decisions sets, this is the first adaptive algorithm for online linear
optimization with a non-vacuous regret bound.

We also study a popular scale-free variant of online mirror descent algorithm,
and we show that in two natural settings it has linear or worse regret.

>>>>

We design algorithms for online linear optimization that have optimal regret
and at the same time do not need to know any upper or lower bounds on the norm
of the loss vectors. We design them as instantiation of the Follow the Regularized Leader (FTRL) and Mirror Descent (MD) meta-algorithms.
We achieve adaptiveness to the norms of the loss vectors
by scale invariance, i.e., our algorithms make exactly the same decisions if
the sequence of loss vectors is multiplied by any positive constant.  The algorithm based on FTRL works for any decision set, bounded or unbounded.  For unbounded decisions sets, this is the first adaptive algorithm for online linear
optimization with a non-vacuous regret bound.
We also show lower bounds on scale-free algorithms based on MD on unbounded domains.

### 2016-09-15 revision/introduction.tex
changes for reviewer 1

\begin{table}[t]
\fontsize{8}{8.2}\selectfont
\centering
\begin{tabular}{|p{3.6cm}|c|p{3.4cm}|c|}
\hline
\textbf{Algorithm} & \textbf{Decisions Set(s)} & \textbf{Regularizer(s)} & \textbf{Scale-Free} \\ \hline \hline
\textsc{Hedge} \cite{Freund-Schapire-1997} & Probability Simplex & Negative Entropy & No \\ \hline
\textsc{GIGA} \cite{Zinkevich-2003} & Any Bounded & $\frac{1}{2}\|w\|_2^2$ & No \\ \hline
\textsc{RDA} \cite{Xiao-2010} & \textbf{Any} & \textbf{Any Strongly Convex} & No \\ \hline
\textsc{FTRL-Proximal} \cite{McMahan-Streeter-2010,McMahan-2014} & Any Bounded & $\frac{1}{2}\|w\|_2^2 + $ any convex func.\footnotemark & \textbf{Yes} \\ \hline
\textsc{AdaGrad MD} \cite{Duchi-Hazan-Singer-2011} & Any Bounded & $\frac{1}{2}\|w\|_2^2 + $ any convex func. & \textbf{Yes} \\ \hline
\textsc{AdaGrad FTRL} \cite{Duchi-Hazan-Singer-2011} & \textbf{Any} & $\frac{1}{2}\|w\|_2^2 + $ any convex func. & No \\ \hline
\textsc{AdaHedge} \cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} & Probability Simplex & Negative Entropy & \textbf{Yes} \\ \hline
\textsc{NAG} \cite{Ross-Mineiro-Langford-2013} & $\{u: \max_t \langle \ell_t, u\rangle \le C\}$ & $\frac{1}{2}\|w\|_2^2 $& Partially\footnotemark \\ \hline
\textsc{Scale invariant algorithms} \cite{Orabona-Crammer-Cesa-Bianchi-2014} & \textbf{Any} & $\frac{1}{2}\|w\|_p^2 + $ any convex func. \newline $1 < p \le 2$ & Partially\textsuperscript{\ref{footnote-label2}} \\ \hline
%\textsc{AdaFTRL} \textbf{[this paper]} & Any Bounded & \textbf{Any Strongly Convex} & \textbf{Yes} \\ \hline
\textsc{Scale-free MD} \textbf{[folklore]} & $\sup_{u,v \in K} \Breg_f(u,v) < \infty$ & \textbf{Any Strongly Convex} & \textbf{Yes} \\ \hline
\textsc{SOLO FTRL} \textbf{[this paper]} & \textbf{Any} & \textbf{Any Strongly Convex} & \textbf{Yes} \\ \hline
\end{tabular}
\caption{Selected results for OLO. Best results in each column are in bold.
\label{table:results}
}
\end{table}

@@

There is a clear practical need to design algorithms that adapt automatically
to the norms of the loss vectors.  A natural, yet overlooked, design method to
achieve this type of adaptivity is by insisting to have a \textbf{scale-free}
algorithm.  That is, the sequence of decisions of the algorithm does not change
if the sequence of loss vectors is multiplied by a positive constant.  The most
important property of scale-free algorithms is that both their loss and their
regret scale linearly with the maximum norm of the loss vector appearing in the
sequence.

@@

The majority of the existing algorithms for OLO are based on two generic algorithms:
\textsc{Follow The Regularizer Leader} (\textsc{FTRL}) and \textsc{Mirror
Descent} (\textsc{MD}). \textsc{FTRL} dates back to the potential-based
forecaster in \cite[Chapter~11]{Cesa-Bianchi-Lugosi-2006} and its theory was
developed in \cite{Shalev-Shwartz-2007}.  The name \textsc{Follow The
Regularized Leader} comes from~\cite{Abernethy-Hazan-Rakhlin-2008}.
Independently, the same algorithm was proposed in~\cite{Nestorov-2009} for
convex optimization under the name \textsc{Dual Averaging} and rediscovered
in~\cite{Xiao-2010} for online convex optimization. The general
version with time-varying regularizers was analyzed in
\cite{Orabona-Crammer-Cesa-Bianchi-2014}. \textsc{MD} was originally proposed
in \cite{Nemirovski-Yudin-1983} and later analyzed in~\cite{Beck-Teboulle-2003}
for convex optimization. In the online learning literature it makes his first
appearance, with a different name, in~\cite{Kivinen-Warmuth-1997}.
%A good description of \textsc{MD} is given in~\cite{Rakhlin-Sridharan-2009}.

@@

We design and analyze three scale-free algorithms: \textsc{AdaFTRL},
\textsc{SOLO FTRL} and \textsc{Scale-Free MD}.  \textsc{AdaFTRL} and
\textsc{SOLO FTRL} are based on \textsc{FTRL}.  \textsc{AdaFTRL} is a
generalization of
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} to arbitrary
strongly convex regularizers.  \textsc{SOLO FTRL} can be viewed as the
``correct'' scale-free version of the diagonal version of \textsc{AdaGrad
FTRL}~\cite{Duchi-Hazan-Singer-2011} generalized to arbitrary strongly convex
regularizers.  \textsc{Scale-Free MD} is based on \textsc{MD}. It is a
generalization of \textsc{AdaGrad MD}~\cite{Duchi-Hazan-Singer-2011} to
arbitrary strongly convex regularizers.  The three algorithms are presented in
Sections~\ref{section:ada-ftrl}, \ref{section:solo-ftrl} and
\ref{section:mirror-descent} respectively.

>>>>

\begin{table}[t]
\fontsize{8}{8.2}\selectfont
\centering
\begin{tabular}{|p{3.6cm}|c|p{3.4cm}|c|}
\hline
\textbf{Algorithm} & \textbf{Decisions Set(s)} & \textbf{Regularizer(s)} & \textbf{Scale-Free} \\ \hline \hline
\textsc{Hedge} \cite{Freund-Schapire-1997} & Probability Simplex & Negative Entropy & No \\ \hline
\textsc{GIGA} \cite{Zinkevich-2003} & Any Bounded & $\frac{1}{2}\|w\|_2^2$ & No \\ \hline
\textsc{RDA} \cite{Xiao-2010} & \textbf{Any} & \textbf{Any Strongly Convex} & No \\ \hline
\textsc{FTRL-Proximal} \cite{McMahan-Streeter-2010,McMahan-2014} & Any Bounded & $\frac{1}{2}\|w\|_2^2 + $ any convex func.\footnotemark & \textbf{Yes} \\ \hline
\textsc{AdaGrad MD} \cite{Duchi-Hazan-Singer-2011} & Any Bounded & $\frac{1}{2}\|w\|_2^2 + $ any convex func. & \textbf{Yes} \\ \hline
\textsc{AdaGrad FTRL} \cite{Duchi-Hazan-Singer-2011} & \textbf{Any} & $\frac{1}{2}\|w\|_2^2 + $ any convex func. & No \\ \hline
\textsc{AdaHedge} \cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} & Probability Simplex & Negative Entropy & \textbf{Yes} \\ \hline
\textsc{NAG} \cite{Ross-Mineiro-Langford-2013} & $\{u: \max_t \langle \ell_t, u\rangle \le C\}$ & $\frac{1}{2}\|w\|_2^2 $& Partially\footnotemark \\ \hline
\textsc{Scale invariant algorithms} \cite{Orabona-Crammer-Cesa-Bianchi-2014} & \textbf{Any} & $\frac{1}{2}\|w\|_p^2 + $ any convex func. \newline $1 < p \le 2$ & Partially\textsuperscript{\ref{footnote-label2}} \\ \hline
%\textsc{AdaFTRL} \textbf{[this paper]} & Any Bounded & \textbf{Any Strongly Convex} & \textbf{Yes} \\ \hline
\textsc{Scale-free MD} \textbf{[this paper]} & $\sup_{u,v \in K} \Breg_f(u,v) < \infty$ & \textbf{Any Strongly Convex} & \textbf{Yes} \\ \hline
\textsc{SOLO FTRL} \textbf{[this paper]} & \textbf{Any} & \textbf{Any Strongly Convex} & \textbf{Yes} \\ \hline
\end{tabular}
\caption{Selected results for OLO. Best results in each column are in bold.
\label{table:results}
}
\end{table}

@@

There is a clear practical need to design algorithms that adapt automatically
to the norms of the loss vectors.  A natural, yet overlooked, design method to
achieve this type of adaptivity is by insisting to have a \textbf{scale-free}
algorithm.  That is, with the same parameters, the sequence of decisions of the algorithm does not change
if the sequence of loss vectors is multiplied by a positive constant.  The most
important property of scale-free algorithms is that both their loss and their
regret scale linearly with the maximum norm of the loss vector appearing in the
sequence.

@@

The majority of the existing algorithms for OLO are based on two generic algorithms:
\textsc{Follow The Regularizer Leader} (\textsc{FTRL}) and \textsc{Mirror
Descent} (\textsc{MD}). \textsc{FTRL} dates back to the potential-based
forecaster in \cite[Chapter~11]{Cesa-Bianchi-Lugosi-2006} and its theory was
developed in \cite{Shalev-Shwartz-2007}.  The name \textsc{Follow The
Regularized Leader} comes from~\cite{Abernethy-Hazan-Rakhlin-2008}.
Independently, the same algorithm was proposed in~\cite{Nestorov-2009} for
convex optimization under the name \textsc{Dual Averaging} and rediscovered
in~\cite{Xiao-2010} for online convex optimization. Time-varying regularizers were analyzed
in~\cite{Duchi-Hazan-Singer-2011} and generalized in \cite{Orabona-Crammer-Cesa-Bianchi-2014}. \textsc{MD} was originally proposed
in \cite{Nemirovski-Yudin-1983} and later analyzed in~\cite{Beck-Teboulle-2003}
for convex optimization. In the online learning literature it makes his first
appearance, with a different name, in~\cite{Kivinen-Warmuth-1997}.
%A good description of \textsc{MD} is given in~\cite{Rakhlin-Sridharan-2009}.

@@

We design and analyze two scale-free algorithms: \textsc{SOLO FTRL} and \textsc{Scale-Free MD}.  A third one, \textsc{AdaFTRL}, is present in the Appendix~\ref{section:ada-ftrl}. \textsc{AdaFTRL} and
\textsc{SOLO FTRL} are based on \textsc{FTRL}.  \textsc{AdaFTRL} is a
generalization of
\textsc{AdaHedge}~\cite{de-Rooij-van-Erven-Grunwald-Koolen-2014} to arbitrary
strongly convex regularizers.  \textsc{SOLO FTRL} can be viewed as the
``correct'' scale-free version of the diagonal version of \textsc{AdaGrad
FTRL}~\cite{Duchi-Hazan-Singer-2011} generalized to arbitrary strongly convex
regularizers.  \textsc{Scale-Free MD} is based on \textsc{MD}. It is a
generalization of \textsc{AdaGrad MD}~\cite{Duchi-Hazan-Singer-2011} to
arbitrary strongly convex regularizers.  The three algorithms are presented in
Sections~\ref{section:ada-ftrl}, \ref{section:solo-ftrl} and
\ref{section:mirror-descent} respectively.

### 2016-09-15 revision/solo-ftrl-proofs.tex
changes for reviewer 1

\begin{proof}[Proof of Lemma~\ref{lemma:sum-of-square-roots-inverses}]
We use the inequality $x/\sqrt{x+y} \le 2(\sqrt{x+y} - \sqrt{y})$ which holds
for non-negative $x,y$ that are not both zero. Substituting $x = a_t$ and $y=\sum_{s=1}^{t-1} a_s$,
we get that for any $t \ge 1$,
$$
\frac{a_t}{\sqrt{\sum_{s=1}^t a_s}} \le 2 \sqrt{\sum_{s=1}^t a_s} \ - \ 2 \sqrt{\sum_{s=1}^{t-1} a_s} \; .
$$
Summing the above inequality over all $t=1,2,\dots,T$, the right side telescopes to
$2 \sqrt{\sum_{t=1}^T a_t}$.
\end{proof}

\begin{proof}[Proof of Corollary~\ref{corollary:regret-solo-ftrl-bounded-set}]
Let $S = \sup_{v \in K} f(v)$. Theorem~\ref{theorem:regret-solo-ftrl} applied
to the regularizer $R(w) = \frac{c}{\sqrt{S}} f(w)$, together with
Proposition~\ref{proposition:diameter-vs-range} and a crude bound
$\max_{t=1,2,\dots,T} \|\ell_t\|_* \le \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2}$,
give
$$
\Regret_T \le \left(c + \frac{2.75}{c}  + 3.5\sqrt{8} \right) \sqrt{S \sum_{t=1}^{T} \norm{\ell_t}_*^2} \; .
$$
We choose $c$ by minimizing $g(c) = c + \frac{2.75}{c} + 3.5\sqrt{8}$. Clearly,
$g(c)$ has minimum at $c = \sqrt{2.75}$ and has minimal value $g(\sqrt{2.75}) =
2\sqrt{2.75} + 3.5\sqrt{8} \le 13.3$.
\end{proof}

>>>>

\begin{proof}[Proof of Lemma~\ref{lemma:sum-of-square-roots-inverses}]
We use the inequality $x/\sqrt{x+y} \le 2(\sqrt{x+y} - \sqrt{y})$ which holds
for non-negative $x,y$ that are not both zero. Substituting $x = a_t$ and $y=\sum_{i=1}^{t-1} a_i$,
we get that for any $t \ge 1$,
$$
\frac{a_t}{\sqrt{\sum_{i=1}^t a_i}} \le 2 \sqrt{\sum_{i=1}^t a_i} \ - \ 2 \sqrt{\sum_{i=1}^{t-1} a_i} \; .
$$
Summing the above inequality over all $t=1,2,\dots,T$, the right side telescopes to
$2 \sqrt{\sum_{t=1}^T a_t}$.
\end{proof}

### 2016-09-15 revision/solo-ftrl.tex
changes for reviewer 1

@@

When $K$ is bounded, we can choose the optimal multiple of the regularizer.  We
choose $R(w) = \lambda f(w)$ where $f$ is a $1$-strongly convex function and
optimize $\lambda$.  The result of the optimization is
Corollary~\ref{corollary:regret-solo-ftrl-bounded-set}; the proof is
in~\ref{section:solo-ftrl-proofs}.
%It is similar to Corollary~\ref{corollary:ada-ftrl-regret-bound} for \textsc{AdaFTRL}. The scaling however is different in the two corollaries.  In Corollary~\ref{corollary:ada-ftrl-regret-bound}, $\lambda \sim 1/(\sup_{v \in K} f(v))$ while in Corollary~\ref{corollary:regret-solo-ftrl-bounded-set} we have $\lambda \sim 1/\sqrt{\sup_{v \in K} f(v)}$.

\begin{corollary}[Regret Bound for Bounded Decision Sets]
\label{corollary:regret-solo-ftrl-bounded-set}
Suppose $K \subseteq V$ is a non-empty bounded closed convex set.  Suppose that
$f:K \to \R$ is a non-negative lower semi-continuous function that is
$1$-strongly convex with respect to $\|\cdot\|$. \textsc{SOLO FTRL} with
regularizer
$$
R(w) = \frac{f(w)\sqrt{2.75}}{\sqrt{\sup_{v \in K} f(v)}}
\quad \text{satisfies} \quad
\Regret_T \le 13.3 \sqrt{\sup_{v \in K} f(v) \sum_{t=1}^{T} \norm{\ell_t}_*^2} \; .
$$
\end{corollary}

@@

\begin{lemma}[Simple Inequality]
\label{lemma:sum-of-square-roots-inverses}
Let $a_1, a_2, \dots, a_T$ be non-negative real numbers. If $a_1 > 0$ then,
$$
\sum_{t=1}^T \frac{a_t}{\sqrt{\sum_{i=1}^t a_i}} \le 2 \sqrt{\sum_{t=1}^T a_t} \; .
$$
\end{lemma}
%
Lemma~\ref{lemma:sum-of-square-roots-inverses} is from
\cite[Lemma~3.5]{Auer-Cesa-Bianchi-Gentile-2002}.  We give a proof
in~\ref{section:solo-ftrl-proofs} for completeness.

>>>>

In this section, we introduces our first scale-free algorithm, based on \textsc{FTRL}.

@@

When $K$ is unbounded, we pay a penalty that scales as $\max_{t \le T} \|\ell_t\|_* \sqrt{T}$, that has the same magnitude of the first term in the bound.
On the other hand, when $K$ is bounded, the second term is a constant and we can choose the optimal multiple of the regularizer.  We
choose $R(w) = \lambda f(w)$ where $f$ is a $1$-strongly convex function and
optimize $\lambda$.  The result of the optimization is
Corollary~\ref{corollary:regret-solo-ftrl-bounded-set}.
%; the proof is in~\ref{section:solo-ftrl-proofs}.
%It is similar to Corollary~\ref{corollary:ada-ftrl-regret-bound} for \textsc{AdaFTRL}. The scaling however is different in the two corollaries.  In Corollary~\ref{corollary:ada-ftrl-regret-bound}, $\lambda \sim 1/(\sup_{v \in K} f(v))$ while in Corollary~\ref{corollary:regret-solo-ftrl-bounded-set} we have $\lambda \sim 1/\sqrt{\sup_{v \in K} f(v)}$.

\begin{corollary}[Regret Bound for Bounded Decision Sets]
\label{corollary:regret-solo-ftrl-bounded-set}
Suppose $K \subseteq V$ is a non-empty bounded closed convex set.  Suppose that
$f:K \to \R$ is a non-negative lower semi-continuous function that is
$1$-strongly convex with respect to $\|\cdot\|$. \textsc{SOLO FTRL} with
regularizer
$$
R(w) = \frac{f(w)\sqrt{2.75}}{\sqrt{\sup_{v \in K} f(v)}}
\quad \text{satisfies} \quad
\Regret_T \le 13.3 \sqrt{\sup_{v \in K} f(v) \sum_{t=1}^{T} \norm{\ell_t}_*^2} \; .
$$
\end{corollary}
%
\begin{proof}
Let $S = \sup_{v \in K} f(v)$. Theorem~\ref{theorem:regret-solo-ftrl} applied
to the regularizer $R(w) = \frac{c}{\sqrt{S}} f(w)$, together with
Proposition~\ref{proposition:diameter-vs-range} and a crude bound
$\max_{t=1,2,\dots,T} \|\ell_t\|_* \le \sqrt{\sum_{t=1}^T \|\ell_t\|_*^2}$,
give
$$
\Regret_T \le \left(c + \frac{2.75}{c}  + 3.5\sqrt{8} \right) \sqrt{S \sum_{t=1}^{T} \norm{\ell_t}_*^2} \; .
$$
We choose $c$ by minimizing $g(c) = c + \frac{2.75}{c} + 3.5\sqrt{8}$. Clearly,
$g(c)$ has minimum at $c = \sqrt{2.75}$ and has minimal value $g(\sqrt{2.75}) =
2\sqrt{2.75} + 3.5\sqrt{8} \le 13.3$.
\end{proof}

@@

\begin{lemma}[{{\cite[Lemma~3.5]{Auer-Cesa-Bianchi-Gentile-2002}}}]
\label{lemma:sum-of-square-roots-inverses}
Let $a_1, a_2, \dots, a_T$ be non-negative real numbers. If $a_1 > 0$ then,
$$
\sum_{t=1}^T \frac{a_t}{\sqrt{\sum_{i=1}^t a_i}} \le 2 \sqrt{\sum_{t=1}^T a_t} \; .
$$
\end{lemma}
%
For completeness, a proof of Lemma~\ref{lemma:sum-of-square-roots-inverses} is in
in~\ref{section:solo-ftrl-proofs}.
